\documentclass[10pt,DIV16,twocolumn,numbers=noenddot]{scrartcl}

\usepackage[utf8x]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amsmath,amssymb}
\usepackage[german]{authblk}
\usepackage[format=plain,labelfont={bf}]{caption}
\usepackage{listings}
\usepackage[shortcuts]{extdash} % \-/
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{multirow}
\setkomafont{disposition}{\bfseries}

\newcommand{\mytitle}{Comparison of methods for the calculation of the real dilogarithm regarding instruction-level parallelism}
\newcommand{\myauthor}{Alexander Voigt}
\newcommand{\Li}{\operatorname{Li}_2}
\newcommand{\secref}[1]{Section~\ref{#1}}
\newcommand{\tabref}[1]{\tablename~\ref{#1}}

% source code highlighting
\lstset{breaklines=true,
  breakatwhitespace=true,
  % numbers=left,
  % stepnumber=1,
  basicstyle=\ttfamily\lst@ifdisplaystyle\footnotesize\fi, % code block in smaller font
  commentstyle=\ttfamily,
  prebreak={\textbackslash},
  breakindent=10pt,
  breakautoindent=false,
  showspaces=false,
  showstringspaces=false,
  frame=single,
  rulesep=0.1em,
  abovecaptionskip=0em,
  aboveskip=1.5em,
  belowcaptionskip=0.5em,
  belowskip=1em,
  keywordstyle=\color{blue},
  stringstyle=\color{red},
  commentstyle=\color{gray},
}

\hypersetup{
  pdftitle   = {\mytitle},
  pdfauthor  = {\myauthor},
  colorlinks = true,
  linkcolor  = blue,
  citecolor  = blue,
  urlcolor   = blue,
  linkcolor  = blue
}

\title{\mytitle}
\author{\myauthor}
\affil{Fachbereich Energie und Biotechnologie, Hochschule Flensburg,\\ Kanzleistra{\ss}e 91--93, 24943 Flensburg, Germany}
\date{\today}

\begin{document}
\maketitle

\section*{Abstract}

We compare different methods for the computation of the real
dilogarithm regarding their ability for using instruction\-/level
parallelism when executed on appropriate CPUs.  As a result we present
an instruction\-/level\-/aware method and compare it to existing
implementations.

\section{Introduction}

The dilogarithm \cite{lewin} is a special function, which appears in
many physics applications, for example in the calculation of quantum
corrections in quantum field theory.  The investigation of new models
beyond the Standard Model of particle physics with a large parameter
space require the evaluation of the dilogarithm (and other functions)
billions of times.  Such investigations thus require a fast evaluation
of the dilogarithm.  In the past many time-efficient algorithms have
been invented (see e.g.\ \cite{koelbigDilog,ginsberg,morris}) and many
are implemented, for example, in general-purpose physics and
mathematics program libraries \cite{root,gsl,cephes}.

Many of the publicly available implementations, e.g.\ \cite{root,gsl},
use formulations for the approximation of the dilogarithm with a
strictly sequential execution of floating\-/point operations.  The
development of modern CPUs with support for instruction\-/level
parallelism (ILP) opens the opportunity for a further reduction of the
run\-/time of the calculation of the dilogarithm.  To make use of ILP,
the algorithm to calculate the dilogarithm must be formulated
appropriately in order to allow for the parallel execution of multiple
floating\-/point operations.  In this paper a method for the
calculation of the real dilogarithm $\Li$ is presented, which makes
use of ILP, potentially resulting in a reduced run\-/time compared to
many algorithms presented so far.

This paper is organized as follows: In \secref{sec:toymodel} different
implementation strategies are discussed using a toy example function.
In \secref{sec:algorithm} an ILP-aware method for the numerical
calculation of the dilogarithm is presented.  A C implementation of
this method can be found in the appendix and in the arXiv submission
of this paper.

\section{Implementation strategies}
\label{sec:toymodel}

As a toy example we consider the function $\ln(1+x)$, which has the
Taylor expansion
%
\begin{align}
  \ln(1+x) \approx x - \frac{x^2}{2} + \frac{x^3}{3} - \frac{x^4}{4} + \frac{x^5}{5} - \frac{x^6}{6} + \frac{x^7}{7}
  \label{eq:taylor}
\end{align}
%
around $x=0$ for $x\in(-1,1]$.  In Eq.~\eqref{eq:taylor} terms of
$O(x^8)$ and beyond have been omitted for brevity.  In the following
we will briefly show various possible implementations of this function
and discuss the performance of the different implementations.  For
brevity we will restrict ourselves to $x\in(-1,1]$.  In C the function
in Eq.~\eqref{eq:taylor} could be implemented very naively as follows:
%
\lstinputlisting[language=C]{src/log1p_naive.c}
%
This naive implementation requires 27 floating\-/point multiplications
and 6 floating\-/point additions.  Although all summands in this
implementation can in principle be calculated in parallel, the cost of
the many (redundant) floating\-/point multiplications is usually so
high that this formulated is in total very time\-/inefficient.

As is well known, the performance of this naive implementation can be
improved by applying Horner's scheme, which reduces the number of
expensive floating\-/point multiplications to 7 and requires 6
additions:
%
\begin{align}
\begin{split}
  \ln(1+x) \approx{}& x \bigg\{1 + x \bigg[-\frac{1}{2} + x \bigg(\frac{1}{3} + x \bigg\{-\frac{1}{4} \\
  &+ x \bigg[\frac{1}{5} + x\bigg(-\frac{1}{6} + \frac{x}{7}\bigg)\bigg]\bigg\}\bigg)\bigg]\bigg\}.
\end{split}\label{eq:horner}%
\end{align}
%
A C implementation using Horner's scheme could read
%
\lstinputlisting[language=C]{src/log1p_horner.c}
%
Note, that the formulation using Horner's scheme requires the
evaluation of the floating\-/point operations in strictly sequential
order.  However, the significant reduction of the number of
floating\-/point multiplications, compared to the naive
implementation, usually results in a shorter run\-/time.

There are several different possibilities to reduce the run\-/time of
Horner's method by rewriting the series expansion of $\ln(1+x)$ such
that more terms can be calculated in parallel, while the number
floating\-/point operations is still kept small.  One possibility is
to split the r.h.s.\ of Eq.~\eqref{eq:horner} into two distinct sums,
one containing the even powers and the other containing the odd powers
of $x$,
%
\begin{align}
\begin{split}
  \ln(1+x) \approx{}& x \left\{1 + y \left[\frac{1}{3} + y
  \left(\frac{1}{5} + \frac{y}{7}\right)\right]\right\} \\
  &- y\left[\frac{1}{2} + y\left(\frac{1}{4} + \frac{y}{6}\right)\right],
\end{split}\label{eq:split}%
\end{align}
%
where $y=x^2$.  This formulation requires 8 floating\-/point
multiplications and 6 additions, i.e.\ one more multiplication than
Horner's scheme.  However, in Eq.~\eqref{eq:split} both sums can then
be calculated in parallel, leading to a potential speed-up if the cost
of the additional multiplication is smaller than the gain by the
parallel execution.  A C implementation could read:
%
\lstinputlisting[language=C]{src/log1p_split.c}
%
Another possibility is to rewrite the r.h.s.\ of Eq.~\eqref{eq:taylor}
using Estrin's scheme \cite{estrin},
%
\begin{align}
\begin{split}
  \ln(1+x) \approx{}& x + \left(\frac{x}{3}-\frac{1}{2}\right)
    y \\ &+\left[\left(\frac{x}{7}-\frac{1}{6}\right)
    y+\frac{x}{5}-\frac{1}{4}\right] z,
\end{split}\label{eq:estrin}%
\end{align}
%
where $y=x^2$ and $z=y^2$.  This form also requires 8 floating\-/point
multiplications and 6 additions, as in Eq.~\eqref{eq:split}.  However,
in Estrin's form more terms can be executed in parallel, leading to a
potential further speed-up.  A C implementation using Estrin's scheme
could read:
%
\lstinputlisting[language=C]{src/log1p_estrin.c}

To exploit ILP further, one could use a rational function
approximation, such as a Padé approximant \cite{pade}, where the
numerator and the denominator can be calculated in parallel.  A Padé
approximant of $\ln(1+x)$ at the same order as Eq.~\eqref{eq:taylor}
reads
%
\begin{align}
  \ln(1+x) &\approx x
  \frac{\frac{320 x^3}{40143}+\frac{44320 x^2}{120429}+\frac{51064
    x}{40143}+1}{\frac{5120 x^3}{40143}+\frac{12320 x^2}{13381}+\frac{23712
    x}{13381}+1}
  \label{eq:pade}
  \\
  &= x
  \frac{\left(\frac{320 x}{40143}+\frac{44320}{120429}\right)y+\frac{51064
    x}{40143}+1}{\left(\frac{5120 x}{40143}+\frac{12320}{13381}\right)y+\frac{23712
    x}{13381}+1},
  \label{eq:mixed}
\end{align}
%
where $y=x^2$.  In Eq.~\eqref{eq:mixed} the numerator and the
denominator have been re-written using Estrin's scheme, allowing for
further ILP.  A C implementation of Eq.~\eqref{eq:mixed} could read:
%
\lstinputlisting[language=C]{src/log1p_mixed.c}
%
Note that the value $y=x^2$ can be re-used in the calculation of both
the numerator and in the denominator.  This ``mixed'' implementation
requires 8 floating\-/point multiplications, 1 floating\-/point
division and 6 additions.  The rational function approximation
\eqref{eq:mixed} may be faster for long numerator/denominator
polynomials, where the cost of the floating\-/point division is
smaller than the gain by the parallel execution of the terms in the
numerator and denominator.


\section{ILP-aware approximation of the real dilogarithm}
\label{sec:algorithm}

In the following an ILP-aware implementation of the real dilogarithm
is presented, which allows for the parallel execution of different
floating\-/point operations.  The real dilogarithm is defined as
%
\begin{align}
  \Li(x) &= -\int_0^x \frac{\ln(1-t)}{t} \, \textnormal{d}t, &
  &x<1.
\end{align}
%
For $|x|<1$ the real dilogarithm has the series representation
%
\begin{align}
  \Li(x) = \sum_{k=1}^\infty \frac{x^k}{k^2}.
  \label{eq:series}
\end{align}
%
Further series representations exist, for example the accelerated
series
%
\begin{align}
  \Li(x) = \sum_{k=0}^\infty \frac{B_k}{(k+1)!} [-\ln(1-x)]^{k+1},
\end{align}
%
where $B_k$ are the Bernoulli numbers.  A series representation in
terms of Chebyshev polynomials of the first kind $T_n(x)$ is given in
\cite{luke},
%
\begin{align}
  \Li(x) = \sum_{k=0}^\infty a_n T_n(2x-1),
  \label{eq:cheby}
\end{align}
%
which is used for example in the ROOT program library \cite{root}.
The r.h.s.\ of Eqs.~\eqref{eq:series}--\eqref{eq:cheby} can be summed,
for example, using Horner's scheme or Clenshaw's algorithm
\cite{clenshaw}.  These summation techniques are, however, purely
sequential and thus do not make use of ILP.  In the following an
approximation of the real dilogarithm is presented, which makes use of
ILP, using a rational function approximation in combination with
Estrin's scheme, similar to the approach used in
Eq.~\eqref{eq:mixed}.

We define the real dilogarithm $\Li(x)$ for all $x\in\mathbb{R}$ as
the real part of its complex continuation.  The real dilogarithm can
be mapped onto the interval $[0,1/2]$ by using the following
identities \cite{lewin}:
%
\begin{align}
  \Li(x) ={}& \ln(1-x)\left[\frac{1}{2}\ln(1-x) - \ln(-x)\right] - \frac{\pi^2}{6}
              \notag \\ &+\Li\left(\frac{1}{1-x}\right), \quad x\leq -1, \\
  \Li(x) ={}& -\Li\left(\frac{x}{x-1}\right) - \frac{1}{2}\ln^2(1-x),
              \notag \\ & -1<x<0, \\
  \Li(x) ={}& -\Li(1-x) + \frac{\pi^2}{6} - \ln(x) \ln(1-x),
              \notag \\ &\frac{1}{2}<x<1, \\
  \Li(x) ={}& \frac{\pi^2}{6} - \ln(x)\left[\ln\left(1-\frac{1}{x}\right) + \frac{1}{2}\ln(x)\right]
              \notag \\ &+ \Li\left(1-\frac{1}{x}\right), \quad 1<x\leq 2, \\
  \Li(x) ={}& -\Li\left(\frac{1}{x}\right) + \frac{\pi^2}{3} - \frac{1}{2} \ln^2(x), \quad x>2,
\end{align}
%
and by using the special values $\Li(-1)=-\pi^2/12$, $\Li(0)=0$,
$\Li(1/2)=\pi^2/12-\ln^2(2)/2$, $\Li(1)=\pi^2/6$ and $\Li(2)=\pi^2/4$.
Within the interval $[0,1/2]$ the real dilogarithm has the series
expansion \eqref{eq:series}, which can be written as
%
\begin{align}
  \Li(x) = x \sum_{k=0}^\infty \frac{x^{k}}{(k+1)^2}.
  \label{eq:series2}
\end{align}
%
We approximate the sum on the r.h.s.\ of Eq.~\eqref{eq:series2} in the
form of a rational minimax approximant on the interval $[0,1/2]$ using
the \texttt{MiniMaxApproximation} from Wolfram/Mathematica
\cite{mathematica},
%
\begin{align}
  \sum_{k=0}^\infty \frac{x^{k}}{(k+1)^2} \approx
  \frac{\sum_{k=0}^5 p_kx^k}{\sum_{k=0}^6 q_kx^k}.
  \label{eq:minimax}
\end{align}
%
The coefficients $p_k$ and $q_k$ are listed in \tabref{tab:coeffs}.
The approximant on the r.h.s.\ of Eq.~\eqref{eq:minimax} has a maximum
relative error of around $5\cdot 10^{-17}$, suited for an evaluation
of $\Li(x)$ with double precision.
%
\begin{table}[t]
  \centering
  \caption{Coefficients of the numerator and denominator polynomials
    for the minimax approximant \eqref{eq:minimax}.}
  \begin{tabular}{lr}
    \toprule
    $p_0$ & $ 0.9999999999999999502\cdot 10^{+0}$ \\
    $p_1$ & $-2.6883926818565423430\cdot 10^{+0}$ \\
    $p_2$ & $ 2.6477222699473109692\cdot 10^{+0}$ \\
    $p_3$ & $-1.1538559607887416355\cdot 10^{+0}$ \\
    $p_4$ & $ 2.0886077795020607837\cdot 10^{-1}$ \\
    $p_5$ & $-1.0859777134152463084\cdot 10^{-2}$ \\
    $q_0$ & $ 1.0000000000000000000\cdot 10^{+0}$ \\
    $q_1$ & $-2.9383926818565635485\cdot 10^{+0}$ \\
    $q_2$ & $ 3.2712093293018635389\cdot 10^{+0}$ \\
    $q_3$ & $-1.7076702173954289421\cdot 10^{+0}$ \\
    $q_4$ & $ 4.1596017228400603836\cdot 10^{-1}$ \\
    $q_5$ & $-3.9801343754084482956\cdot 10^{-2}$ \\
    $q_6$ & $ 8.2743668974466659035\cdot 10^{-4}$ \\
    \bottomrule
  \end{tabular}
  \label{tab:coeffs}
\end{table}
%
Note that the approximant on the r.h.s.\ of Eq.~\eqref{eq:minimax} is
very similar to the results presented in Ref.~\cite{morris} (DILOG
0011), although with slightly different coefficients.  In order to
allow for the use of ILP for the evaluation of the numerator and the
denominator we use Estrin's scheme,
%
\begin{align}
  \Li(x) &\approx x\frac{P(x)}{Q(x)}, & &0\leq x \leq \frac{1}{2},
  \label{eq:final}
\end{align}
%
with
%
\begin{align}
  P(x) &= p_0 + x p_1 + y (p_2 + x p_3) + z (p_4 + x p_5), \\
  Q(x) &= q_0 + x q_1 + y (q_2 + x q_3) + z (q_4 + x q_5 + y q_6),
\end{align}
%
where $y=x^2$ and $z=y^2$.  A C implementation of Eq.~\eqref{eq:final}
can be found in the appendix.

\section{Run\-/time comparison}

In \tabref{tab:runtime} the run\-/time of different C implementations is
compared to the implementation given in the appendix on two different
CPU architectures.  On each architecture the implementations are
compiled with \texttt{g++} with optimization \texttt{-O2}.  The shown
run\-/time denotes the total wall clock time for the calculation of
$\Li(x)$ for $10^6$ uniformly distributed random values $x\in[0,1/2]$.
The choice of this interval ensures that the measurement is only
sensitive to the formulation of the approximant, but not to the
transformation onto the interval $[0,1/2]$.%
%
\begin{table}[tb]
  \centering
  \caption{Run\-/time in seconds for the calculation of $\Li(x)$ for
    different implementations on different CPU architectures with
    different compilers.}
  \begin{tabular}{lcc}
    \toprule
    \multirow{2}{*}{Implementation} & i7-4700MQ & i7-5600U \\
    & \texttt{g++} 9.3.0 & \texttt{g++} 10.2.1 \\
    \midrule
    Eq.~\eqref{eq:final} (this paper) & $0.0076$ & $0.0058$ \\
    Morris (DILOG 0011) \cite{morris} & $0.0115$ & $0.0085$ \\
    Cephes \cite{cephes}              & $0.0136$ & $0.0119$ \\
    Algorithm 327 \cite{koelbigDilog} & $0.0229$ & $0.0172$ \\
    Chebyshev \cite{luke,root}        & $0.0718$ & $0.0515$ \\
    Algorithm 490 \cite{ginsberg}     & $0.1057$ & $0.0576$ \\
    GSL \cite{gsl}                    & $0.2091$ & $0.0898$ \\
    \bottomrule
  \end{tabular}
  \label{tab:runtime}
\end{table}%

We find that the implementation presented in this paper and the
approach presented in Ref.~\cite{morris} (DILOG 0011) are fastest
among the ones shown in the table.  Both approaches use a rational
function approximant with the same number of terms.  However, the
approximant from Ref.~\cite{morris} has been implemented (for
comparison) to use Horner's scheme to evaluate the numerator and the
denominator.  Thus, the run\-/time difference of $\sim 30\%$ between
the two stems only from the fact that the implementation presented
in this paper uses Estrin's scheme to evaluate the numerator and the
denominator and as such can make use of ILP, while in the
implementation of Ref.~\cite{morris} the numerator and the denominator
must be evaluated sequentially.

The third fastest implementation is provided by Cephes \cite{cephes},
which implements a rational function approximation for Spence's
function $\operatorname{Sp}(x)=\Li(1-x)$.  The implementation provided
by Cephes is by a factor two slower than the implementation presented
in this paper.  This longer run\-/time has several reasons: For one,
Cephes uses a rational function approximation with more terms in the
numerator and denominator.  Furthermore, Cephes evaluates the
numerator and denominator polynomials using Horner's scheme, which is
purely sequential.\footnote{For $x>1$ Cephes calculates $\ln(x)$
  twice, which leads to a further increased run\-/time when a benchmark
  interval with $x>1$ is chosen.}

Algorithm 327 \cite{koelbigDilog}, Algorithm 490 \cite{ginsberg} and
the implementation found in the GNU Scientific Library (GSL)
\cite{gsl} use series expansions, which are summed using Horner's
scheme.  These algorithms thus do not make use of ILP, which is one
reason for their longer run\-/time.  Furthermore, Algorithm 490 and most
series expansions in the GSL use floating\-/point divisions inside
loops, which may be another reason for the increased run\-/time.

A representation in terms of Chebyshev polynomials \cite{luke} is
implemented in the ROOT program library \cite{root}.  In this
implementation the polynomials are summed using Clenshaw's algorithm,
which is also purely sequential and slightly more complex than
Horner's scheme.  This is a reason why the implementation in
\cite{root} has a larger run\-/time than the rational function
approximants discussed above.

\section{Summary}

In this paper different approaches for the numerical calculation of
the real dilogarithm have been compared w.r.t.\ their potential for
the use of instruction\-/level (ILP) parallelism.  We found that a
rational minimax approximant in combination with Estrin's scheme to
evaluate the numerator/denominator polynomials provides both a
sufficient numerical precision and a high degree of ILP, potentially
leading to a reduced run\-/time.  An approximant and a corresponding C
implementation have been proposed and compared to several existing
implementations w.r.t.\ their run\-/time.  The C implementation of the
presented approximant can be found in the appendix.  C++ and FORTRAN
implementations can be found in the Polylogarithm library
\cite{polylogarithm}.  Implementations in Julia and Rust can be found
in \cite{PolyLog.jl} and \cite{polylog}, respectively.

\appendix

\section{Implementation of the real dilogarithm}

\lstinputlisting[language=C]{anc/Li2.c}

\bibliographystyle{ieeetr}
\bibliography{paper}

\end{document}
